---
title: "derive_vars"
author: "Michael Leukam"
date: "2021-02-03"
output: 
  workflowr::wflow_html
  html_notebook
editor_options:
  chunk_output_type: console
---
## Setup
Clear the workspace and set seed
```{r}
rm(list=ls())
set.seed(818)
```


Read in packages
```{r}
# Bioconductor
library(Biobase)
library(vegan)

# CRAN
library(Matrix.utils)
library(tidyverse)
library(ggpubr)
```

```{r}
knitr::opts_chunk$set(eval = FALSE)
```

Read in data
From tcga_data_prep
```{r}
# tcga_counts <- readRDS("data/comb_tcga_counts_list.rds")

# mode 2 data
# tcga_counts <- readRDS("data/mode2_tcga_counts_list.rds")

# str(tcga_counts, max.level = 2)
# genus_tbl <- tcga_counts$counts$genus$STAD %>%
#  print()
# imm_cor_tbl <- tcga_counts$imm_cor$STAD %>%
#   print()

# batch corrected and uncorrected data
genus_counts_es_list <- readRDS("output/genus_combined_expressionsets.rds")
str(genus_counts_es_list, max.level = 1)

genus_counts_es_list

# function for converting expression set into list of tumors
bytumortyper <- function(es){
  clindat <- pData(es) %>% 
    rownames_to_column(var = "tcga_barcode") %>%
    as_tibble()
  proj_lookup <- clindat %>%
    select(tcga_barcode, tcga_proj)
  counts <- exprs(es) %>%
    t() %>%
    print() %>%
    as.data.frame() %>%
    rownames_to_column(var = "tcga_barcode") %>%
    as_tibble() %>%
    left_join(proj_lookup) %>%
    select(tcga_proj, tcga_barcode, everything()) %>%
    group_by(tcga_proj) %>%
    nest() %>%
    deframe()
  list("counts" = counts, "clindat" = clindat)
}

# test
corrected_cpm <- bytumortyper(genus_counts_es_list$cbseq_corrected_counts)

# apply to list of ES
counts_data <- map(genus_counts_es_list, bytumortyper)
str(counts_data, max.level = 2)

```

```{bash}
sshfs mjl62@o2.hms.harvard.edu:/n/scratch3/users/m/mw292/AGAMEMNON/TCGA/Results/ ~/sshfs_mount/TCGA

cd ~/sshfs_mount/TCGA/TCGA_STAD/7c564146-979c-4b3d-9a0d-b9a1e0d07ca2/Analysis/7c564146-979c-4b3d-9a0d-b9a1e0d07ca2/taxonomic_ranks
```

Read in taxonomy data
```{r}
taxonomy <- read_csv("data/tcga_proj_taxonomy.csv")
taxonomy_nomissing <- readRDS("data/tcga_proj_taxonomy_nomissing.rds")
```

Goal:
% microbial vs total reads
% microbial vs uniquely mapped on human reads
% viral (as above)
bacteria vs microbes
Number of different species
Genera, species, strains per million reads mapped on human
Genera, species, strain diversity and distribution descriptors (Shannon, Simpson, etc)
Genera abundance, species abundance, and strain abundance as correlates

## Descriptive Statistics

#### Distribution

Visualize distributions
```{r}
# make tidy dataframes for plotting
# run this over the entire counts data object
tidy_generator <- function(listelement){
  listelement$counts %>%
    map(., function(df){
      df %>% 
        pivot_longer(-tcga_barcode, 
          names_to = "genus", 
          values_to = "counts") %>%
        arrange(desc(counts))})
}

# make tidy frames with cpm
tidy_counts <- map(counts_data, tidy_generator)

str(tidy_counts, max.level = 1)
str(tidy_counts, max.level = 2)

# counts distribution
counts_dist_plotter <- function(element){
  map(element, function(df){
    df %>%
      ggplot(aes(x = counts, color = tcga_barcode)) +
      geom_density() +
      theme(legend.position = "none")}) %>%
  ggarrange(plotlist = .) %>%
  annotate_figure(., 
    top = text_grob(name(listelement), 
    color = "black", 
    face = "bold", 
    size = 14))
}

dist_plot_list <- map(tidy_counts, counts_dist_plotter)

str(dist_plot_list, max.level = 1)
# plot raw distributions for each tumor type

dist_plot_list[["corrected_cpm"]]
dist_plot_list[["uncorrected_cpm"]]
dist_plot_list[["corrected_counts"]]
dist_plot_list[["uncorrected_counts"]]
dist_plot_list[["corrected_microcpm"]]
dist_plot_list[["uncorrected_microcpm"]]

# do.call(ggarrange, dist_plot_list_genus_cpm_corrected[["corrected_cpm"]]) %>%
#  annotate_figure(., top = text_grob("Corrected Genus CPM by Tumor Type", color = "black", face = "bold", size = 14))
          
# log cpm distribution
log_dist_plotter <- function(df){
  plot <- df %>%
    mutate(lcounts = log(counts)) %>%
    dplyr::filter(lcounts > -100) %>%
    ggplot(aes(x = lcounts), color = tcga_barcode) +
      geom_density() +
      theme(legend.position = "none") +
      expand_limits(x = c(4, 5))
}
# log cpm density plots
log_plot_list_genus_cpm_corrected <- map(tidy_genus_cpm_corrected, log_dist_plotter)
names(log_plot_list_genus_cpm_corrected) <- names(tidy_genus_cpm_corrected)
do.call(ggarrange, log_plot_list_genus_cpm_corrected) %>%
  annotate_figure(., top = text_grob("Corrected log(CPM) by Tumor Type", color = "black", face = "bold", size = 14))
```

## Derive new variables

#### CPM alpha diversity

```{r}

# get measures of alpha-diversity
alpha_getter <- function(df){
# get global basic diversity indices
  shannon <- df %>% 
    diversity(., index = "shannon") %>%
    enframe() %>%
    rename(shannon = value)
  simpson <- df %>%
    diversity(., index = "simpson") %>%
    enframe() %>%
    rename(simpson = value)
  invsimpson <- df %>%
    diversity(., index = "invsimpson") %>%
    enframe() %>%
    rename(invsimpson = value)
  # get global pielou evenness measure
  shanvect <- df %>%
    diversity(., index = "shannon")
  pielou <- tibble::enframe(shanvect/log(specnumber(df))) %>%
    rename(pielou_evenness = value)
  # get global rarefied richness
  df_round <- df %>%
    as.matrix() %>%
    round(., digits = 0)
  rar <- rarefy(df_round, min(rowSums(df_round), na.rm = T)) %>%
    enframe() %>%
    rename(rar_rich = value)
# get separate diversity indices for each superkingdom
  # get vectors of genera belonging to each superkingdom
  viral_genera <- taxonomy_nomissing %>%
    dplyr::filter(Superkingdom == "Virus") %>%
    pull(Genus)
  bacterial_genera <- taxonomy_nomissing %>%
    dplyr::filter(Superkingdom == "Bacteria") %>%
    pull(Genus)
  archaea_genera <- taxonomy_nomissing %>%
    dplyr::filter(Superkingdom == "Archaea") %>%
    pull(Genus)
  eukaryota_genera <- taxonomy_nomissing %>%
    dplyr::filter(Superkingdom == "Eukaryota") %>%
    pull(Genus)
  # compute separate diversity metrics
  # virus
  viral_df <- df %>% 
    dplyr::select(any_of(viral_genera))
  viral_shannon <- viral_df %>% 
    diversity(., index = "shannon") %>%
    enframe() %>%
    rename(viral_shannon = value)
  viral_simpson <- viral_df %>% 
    diversity(., index = "simpson") %>%
    enframe() %>%
    rename(viral_shannon = value)
  viral_invsimpson <- viral_df %>% 
    dplyr::select(any_of(viral_genera)) %>%
    diversity(., index = "invsimpson") %>%
    enframe() %>%
    rename(viral_shannon = value)
  viral_shanvect <- viral_df %>%
    diversity(., index = "shannon")
  viral_pielou <- enframe(viral_shanvect/log(specnumber(viral_df))) %>%
    rename(viral_pielou_evenness = value)
  viral_df_round <- viral_df %>%
    as.matrix() %>%
    round(., digits = 0)
  viral_rar <- rarefy(viral_df_round, min(rowSums(viral_df_round), na.rm = T)) %>%
    enframe() %>%
    rename(viral_rar_rich = value)
  # bacterial
  bacterial_df <- df %>% 
    dplyr::select(any_of(bacterial_genera))
  bacterial_shannon <- bacterial_df %>% 
    diversity(., index = "shannon") %>%
    enframe() %>%
    rename(bacterial_shannon = value)
  bacterial_simpson <- bacterial_df %>% 
    diversity(., index = "simpson") %>%
    enframe() %>%
    rename(bacterial_shannon = value)
  bacterial_invsimpson <- bacterial_df %>% 
    dplyr::select(any_of(bacterial_genera)) %>%
    diversity(., index = "invsimpson") %>%
    enframe() %>%
    rename(bacterial_shannon = value)
  bacterial_shanvect <- bacterial_df %>%
    diversity(., index = "shannon")
  bacterial_pielou <- enframe(bacterial_shanvect/log(specnumber(bacterial_df))) %>%
    rename(bacterial_pielou_evenness = value)
  bacterial_df_round <- bacterial_df %>%
    as.matrix() %>%
    round(., digits = 0)
  bacterial_rar <- rarefy(bacterial_df_round, min(rowSums(bacterial_df_round), na.rm = T)) %>%
    enframe() %>%
    rename(bacterial_rar_rich = value)
  # archaea
  archaea_df <- df %>% 
    dplyr::select(any_of(archaea_genera))
  archaea_shannon <- archaea_df %>% 
    diversity(., index = "shannon") %>%
    enframe() %>%
    rename(archaea_shannon = value)
  archaea_simpson <- archaea_df %>% 
    diversity(., index = "simpson") %>%
    enframe() %>%
    rename(archaea_shannon = value)
  archaea_invsimpson <- archaea_df %>% 
    dplyr::select(any_of(archaea_genera)) %>%
    diversity(., index = "invsimpson") %>%
    enframe() %>%
    rename(archaea_shannon = value)
  archaea_shanvect <- archaea_df %>%
    diversity(., index = "shannon")
  archaea_pielou <- enframe(archaea_shanvect/log(specnumber(archaea_df))) %>%
    rename(archaea_pielou_evenness = value)
  archaea_df_round <- archaea_df %>%
    as.matrix() %>%
    round(., digits = 0)
  archaea_rar <- rarefy(archaea_df_round, min(rowSums(archaea_df_round), na.rm = T)) %>%
    enframe() %>%
    rename(archaea_rar_rich = value)
  # eukaryota
  eukaryota_df <- df %>% 
    dplyr::select(any_of(eukaryota_genera))
  eukaryota_shannon <- eukaryota_df %>% 
    diversity(., index = "shannon") %>%
    enframe() %>%
    rename(eukaryota_shannon = value)
  eukaryota_simpson <- eukaryota_df %>% 
    diversity(., index = "simpson") %>%
    enframe() %>%
    rename(eukaryota_shannon = value)
  eukaryota_invsimpson <- eukaryota_df %>% 
    dplyr::select(any_of(eukaryota_genera)) %>%
    diversity(., index = "invsimpson") %>%
    enframe() %>%
    rename(eukaryota_shannon = value)
  eukaryota_shanvect <- eukaryota_df %>%
    diversity(., index = "shannon")
  eukaryota_pielou <- tibble::enframe(eukaryota_shanvect/log(specnumber(eukaryota_df))) %>%
    rename(eukaryota_pielou_evenness = value)
  eukaryota_df_round <- eukaryota_df %>%
    as.matrix() %>%
    round(., digits = 0)
  eukaryota_rar <- rarefy(eukaryota_df_round, min(rowSums(eukaryota_df_round), na.rm = T)) %>%
    enframe() %>%
    rename(eukaryota_rar_rich = value)
# join columns in single tbl
  alpha_df <- shannon %>%
    left_join(simpson, by = "name") %>%
    left_join(invsimpson, by = "name") %>%
    left_join(pielou, by = "name") %>%
    left_join(rar, by = "name") %>%
    left_join(viral_shannon, by = "name") %>%
    left_join(viral_simpson, by = "name") %>%
    left_join(viral_invsimpson, by = "name") %>%
    left_join(viral_pielou, by = "name") %>%
    left_join(viral_rar, by = "name") %>%
    left_join(bacterial_shannon, by = "name") %>%
    left_join(bacterial_simpson, by = "name") %>%
    left_join(bacterial_invsimpson, by = "name") %>%
    left_join(bacterial_pielou, by = "name") %>%
    left_join(bacterial_rar, by = "name") %>%
    left_join(archaea_shannon, by = "name") %>%
    left_join(archaea_simpson, by = "name") %>%
    left_join(archaea_invsimpson, by = "name") %>%
    left_join(archaea_pielou, by = "name") %>%
    left_join(archaea_rar, by = "name") %>%
    left_join(eukaryota_shannon, by = "name") %>%
    left_join(eukaryota_simpson, by = "name") %>%
    left_join(eukaryota_invsimpson, by = "name") %>%
    left_join(eukaryota_pielou, by = "name") %>%
    left_join(eukaryota_rar, by = "name") %>%
    dplyr::rename(tcga_barcode = name)
  alpha_df
}

# prepare data------------------------------------------
# get counts in dataframe

str(counts_data, max.level = 2)
str(counts_data$cbseq_corrected_counts, max.level = 2)
counts_data$cbseq_corrected_counts$counts$BRCA

make_counts_df <- function(df){
  df %>%
    # add up any duplicate rows to summarize at genus/sample level
    dplyr::group_by(tcga_barcode) %>%
    summarise_all(sum) %>%
    # remove non-numeric values
    filter(across(everything(), ~ !is.nan(.x))) %>%
    # eliminate negative values
    mutate(across(everything(), ~ ifelse(.x < 0, 0, .x))) %>%
    # convert to dataframe with rownames
    as.data.frame() %>%
    column_to_rownames(var = "tcga_barcode")
}

# apply df maker function to each of the data types
cbseq_corrected_counts_df_list <- map(
  counts_data$cbseq_corrected_counts$counts, 
  make_counts_df)

uncorrected_counts_df_list <- map(
  counts_data$uncorrected_counts$counts, 
  make_counts_df)

limma_corrected_counts_df_list <- map(
  counts_data$limma_corrected_counts$counts, 
  make_counts_df)

# apply function to get alpha diversity
cbseq_corrected_counts_alpha <- map(
  cbseq_corrected_counts_df_list, 
  alpha_getter)
uncorrected_counts_alpha <- map(
  uncorrected_counts_df_list, 
  alpha_getter)
limma_corrected_counts_alpha <- map(
  limma_corrected_counts_df_list, 
  alpha_getter)
```


```{r}
# plots
cbseq_corrected_counts_alpha %>%
  enframe() %>%
  unnest(value) %>%
  rename(tcga_proj = name) %>%
  ggplot(aes(x = tcga_proj, y = shannon, fill = shannon)) +
    geom_violin() +
  expand_limits(x = 0, y = 0)

cbseq_corrected_counts_alpha %>%
  enframe() %>%
  unnest(value) %>%
  rename(tcga_proj = name) %>%
  ggplot(aes(x = tcga_proj, y = simpson, fill = simpson)) +
    geom_violin() +
  expand_limits(x = 0, y = 0)

```

#### High-level taxonomy

This block is now defunct
```{r}
# read in repaired toxonomy
taxonomy_nomissing <- readRDS("data/tcga_proj_taxonomy_nomissing.rds")
# problem - Tissierella_Class is still called ---_Class
taxonomy_nomissing <- taxonomy_nomissing %>% 
  mutate(Genus = ifelse(Genus == 	"---_Class", "Tissierella_Class", Genus)) %>%
  arrange(Genus) %>%
  print()

# repair counts not generated with repaired taxonomy
kingdom_list_element <- tcga_counts$cpm_counts$kingdom_cpm$ACC

high_leveler <- function(kingdom_list_element){
  fractional_kingdom <- kingdom_list_element %>%
    dplyr::rename(viral = `---`, archaea = Archaea, 
                  bacteria = Bacteria, 
                  eukaryota = Eukaryota) %>%
    mutate(total_cpm = viral + archaea + bacteria + eukaryota) %>%
    mutate(across(-c(tcga_barcode, total_cpm), ~ . / total_cpm, .names = "frac_{.col}")) %>%
    dplyr::rename(microbe_cpm = total_cpm)
}

# apply function
kingdom_cpm_list <- tcga_counts$counts$kingdom
high_level_vars <- map(kingdom_list, high_leveler)
names(high_level_vars) <- names(kingdom_list)
high_level_vars
```

Start with this block 
```{r}
# get high level taxonomy data for corrected genus CPM
kingdom_lookup <- taxonomy_nomissing %>%
  dplyr::select(Superkingdom, Genus) %>%
  # repair Tissierella_Class
  mutate(Genus = ifelse(Genus == "---_Class", "Tissierella_Class", Genus)) %>%
  print()

# test variable
tcga_list <- cbseq_corrected_counts_df_list

cbseq_corrected_counts_df_list[[1]][1:10, 1:3]
uncorrected_counts_df_list[[1]][1:10, 1:3]
limma_corrected_counts_df_list[[1]][1:10, 1:3]

# function that takes a list of dataframes as input
# where each dataframe is a single tumor type
higher_leveler <- function(tcga_list){
  tcga_list %>%
    # repair the data
    map(~ as.matrix(.)) %>%
    map(~ t(.)) %>%
    map(~ as.data.frame(.)) %>%
    map(~ rownames_to_column(., var = "Genus")) %>%
    map(~ as_tibble(.)) %>%
    # map(~ ) %>%
    map(~ left_join(., kingdom_lookup, by = "Genus")) %>%
    map(~ select(., Superkingdom, everything())) %>%
    map(~ select(., -Genus)) %>%
    # check for NAs or other weird results
    map(~ arrange(., Superkingdom)) %>%
    map(~ group_by(., Superkingdom)) %>%
    map(~ summarise_all(., sum)) %>%
    map(~ as.data.frame(.)) %>%
    map(~ column_to_rownames(., var = "Superkingdom")) %>%
    map(~ as.matrix(.)) %>%
    map(~ t(.)) %>%
    map(~ as.data.frame(.)) %>%
    map(~ rownames_to_column(., var = "tcga_barcode")) %>%
    map(~ as_tibble(.)) %>%
    map(~ select(., 
      tcga_barcode, 
      bacteria = Bacteria,
      viruses = Virus,
      archaea = Archaea,
      eukaryota = Eukaryota)) %>%
    map(~ mutate(., microbial_cpm = viruses + archaea + bacteria + eukaryota)) %>%
    map(~ mutate(., 
      across(-c(tcga_barcode, microbial_cpm), 
      ~ . / microbial_cpm, 
      .names = "frac_{.col}")))
}

hilevel_cbseq_corrected_counts <- higher_leveler(cbseq_corrected_counts_df_list)

hilevel_cbseq_uncorrected_counts <- higher_leveler(uncorrected_counts_df_list)

hilevel_limma_corrected_counts <- higher_leveler(limma_corrected_counts_df_list)
```

## CAN START HERE NEEDS REPAIR Write out results
```{r}
genus_analysis_list <- readRDS(genus_analysis_list, "data/genus_analysis_list.rds")

imm_cor <- read_delim("~/Projects/tcga_micro/data/imm_cor_souce.tsv", delim = "\t")

# limma corrected counts
limma_hilevel_flat <- hilevel_limma_corrected_counts %>%
  bind_rows() %>%
  print()

limma_alpha_flat <- limma_corrected_counts_alpha %>%
  bind_rows() %%
  print()

pData(genus_counts_es_list$limma_corrected_counts) %>%
  rownames_to_column(var = "tcga_barcode") %>%
  as_tibble() %>%
  left_join(
    
    )
  left_join(limma_corrected_counts_alpha, by = "tcga_barcode")
  

# cbseq corrected counts
genus_counts_es_list$cbseq_corrected_counts
  
# uncorrected counts
genus_counts_es_list$uncorrected_counts

genus_analysis_list <- list(
  "cbseq_corrected_high_level_vars" = hilevel_cbseq_corrected_counts,
  "limma_corrected_high_level_vars" = hilevel_limma_corrected_counts,
  "uncorrected_high_level_vars" = uncorrected_counts_df_list,
  "cbseq_corrected_alpha" = cbseq_corrected_counts_alpha,
  "limma_corrected_alpha" = limma_corrected_counts_alpha,
  "uncorrected_alpha" = uncorrected_counts_alpha)
saveRDS(genus_analysis_list, "data/genus_analysis_list.rds")

```


#### Beta diversity
```{r}


```
